Best Practices for Single Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-12-01 paper/manuscript.tex
Tweaks from @sheenam1509

The chemical potential of a solute $u$ in an ideally diluted solution in a solvent $v$ can be written as~\cite{guggenheim1952mixtures504,gilson1997statisticalthermodynamic,gilson2007calculation}
\begin{equation}
  \mu_u = -k_B T \ln \frac{q'_u}{C_u \Lambda_u^3}
  \label{eq:chemical-potential-definition}
\end{equation}
where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal De-Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass. The quantity $q'_u$ in Eq.~(\ref{eq:chemical-potential-definition}) is the molecular partition function of the solute at rest in the solvent
\begin{equation}
  q'_u = \frac{\Lambda_u^3}{\prod_i \lambda_i^3} 8 \pi^2 z_u
  \label{eq:intra-qpu-def}
\end{equation}
where $\beta = 1/(k_B T)$, $\lambda_i$ is the thermal De Broglie wavelength of the $i$-th atom of the solute, and $z_u$ is the intramolecular configurational partition function of the solute
\begin{equation}
  z_u = \int dx_u e^{-\beta \Psi_v(x_u)}
  \label{eq:intra-zu-def}
\end{equation}
where $x_u$ represents the collection of internal degrees of freedom of the solute (bond lengths, bond angles, and dihedral angles) with the corresponding volume element $dx_u$ assumed to include the appropriate Jacobian factors. The function $\Psi_v(x_u)$ in Eq.~\ref{eq:intra-zu-def} is the solvent-averaged potential of mean force of the solute in conformation $x_u$ in solvent $v$\cite{rouxandsimonson1999solventmodels} 
\begin{equation}
  e^{-\beta \Psi_v(x_u)} = \frac{1}{Z_{N_v}} \int d\vec{r}_v e^{-\beta U(x_u, \vec{r}_v)}
  \label{eq:solvent-pmf-def}
\end{equation}
where $\vec{r}_v$ represents the collection of Cartesian coordinates of the atoms of $N_v$ solvent molecules, $U(x_u, \vec{r}_v)$ is the potential energy of the solvent with one solute molecule in conformation $x_u$ placed in an arbitrary position in a volume $V$. When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_u(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$.\cite{gilson2007calculation} However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general. Finally, $Z_{N_v}$ is the configurational partition function of the pure solvent
\begin{equation}
  Z_{N_v} = \int d\vec{r}_v e^{-\beta U(\vec{r}_v)} \, .
  \label{eq:Z-pure-solvent}
\end{equation}

@@

The standard chemical potential of a solute in an ideally dilute solution is obtained from Eq.~(\ref{eq:chemical-potential-definition}) by setting the concentration to the standard concentration $C^\circ$, generally $1$ mol/L equivalent to $1/1,661$ particles/\AA$^{3}$, 
\begin{equation}
  \mu_u^\circ = -k_B T \ln \frac{q'_u}{C^\circ \Lambda_u^3} \, .
  \label{eq:chemical-potential-standard}
\end{equation}
The standard state state of a dilute solution is defined as an ideally dilute solution at concentration $C^\circ$.~\cite{levine2009physicalchemistrybook6ed} Hence, Eq.~(\ref{eq:chemical-potential-standard}) is formally correct for real solutions as well.

@@

To derive an RBFE formulation from double-decoupling, we write expressions for the binding constants of $A$ and $B$ using Eq.~(\ref{eq:ddm-ratios-1}) and taking their ratio. While doing so, the standard state, pure solvent, and gas-phase terms cancel, and we obtain
\begin{equation}\label{eq:double-transformation-Kb-rbfe}
\frac{K_b(B)}{K_b(A)} = \frac{Z_{N_v,RB}}{Z_{N_v,RA}} \frac{Z_{N_v,A}}{Z_{N_v,B}}
\end{equation}
where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution. Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor (Figure \ref{fig:fig_binding_thermodynamic_cycle}):
\begin{equation}\label{eq:double-transformation-dg-rbfe}
\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,A}}{Z_{N_v,B}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
\end{equation}
In practice, each ratio of partition functions must first be turned into an ensemble average to evaluate it. Similar to the preceding case, this step requires manipulating the integrals so that they are carried over the same set of degrees of freedom. As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies capable of describing both ligands. For example, among other features, alchemical topologies include atoms common to both ligands and dummy atoms that exist in one but not in the other ligand (Figure \ref{fig:fig_topology}). In one limit, called dual coordinate dual topology in Section \ref{sec:relative-fe-protocol}, all atoms of both ligands are present, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system. A key aspect of constructing a suitable alchemical topology is the requirement that the additional degrees of freedom do not affect the RBFE estimate.

>>>>

The chemical potential of a solute $u$ in an ideally diluted solution in a solvent $v$ can be written as~\cite{guggenheim1952mixtures504,gilson1997statisticalthermodynamic,gilson2007calculation}
\begin{equation}
  \mu_u = -k_B T \ln \frac{q'_u}{C_u \Lambda_u^3}
  \label{eq:chemical-potential-definition}
\end{equation}
where $k_B$ is Boltzmann's constant, $T$ is the absolute temperature, $C_u = N_u/V$ is the number concentration of the solute, $\Lambda_u = h/\sqrt{2 \pi M_u k_B T}$ is the thermal De Broglie wavelength of the solute, where $h$ is Planck's constant and $M_u$ is the solute's total mass. The quantity $q'_u$ in Eq.~(\ref{eq:chemical-potential-definition}) is the molecular partition function of the solute at rest in the solvent
\begin{equation}
  q'_u = \frac{\Lambda_u^3}{\prod_i \lambda_i^3} 8 \pi^2 z_u
  \label{eq:intra-qpu-def}
\end{equation}
where $\beta = 1/(k_B T)$, $\lambda_i$ is the thermal De Broglie wavelength of the $i$-th atom of the solute, and $z_u$ is the intramolecular configurational partition function of the solute
\begin{equation}
  z_u = \int dx_u e^{-\beta \Psi_v(x_u)}
  \label{eq:intra-zu-def}
\end{equation}
where $x_u$ represents the collection of internal degrees of freedom of the solute (bond lengths, bond angles, and dihedral angles) with the corresponding volume element $dx_u$ assumed to include the appropriate Jacobian factors. The function $\Psi_v(x_u)$ in Eq.~\ref{eq:intra-zu-def} is the solvent-averaged potential of mean force of the solute in conformation $x_u$ in solvent $v$\cite{rouxandsimonson1999solventmodels} 
\begin{equation}
  e^{-\beta \Psi_v(x_u)} = \frac{1}{Z_{N_v}} \int d\vec{r}_v e^{-\beta U(x_u, \vec{r}_v)}
  \label{eq:solvent-pmf-def}
\end{equation}
where $\vec{r}_v$ represents the collection of Cartesian coordinates of the atoms of $N_v$ solvent molecules, $U(x_u, \vec{r}_v)$ is the potential energy of the solvent with one solute molecule in conformation $x_u$ placed in an arbitrary position in a volume $V$. When the potential energy function of the solution is separable into an intramolecular potential energy of the solute $U(x_u)$ and a solute-solvent non-bonded interaction energy, the potential of mean force can be written as the sum $\Psi_v(x_u) = U(x_u) + W_v(x_u)$ where $W_u(x_u)$ is the solvent potential of mean force, the solvation free energy of the solute fixed in configuration $x_u$.\cite{gilson2007calculation} However, the potential of mean force definition Eq.~(\ref{eq:solvent-pmf-def}) is valid in general. Finally, $Z_{N_v}$ is the configurational partition function of the pure solvent
\begin{equation}
  Z_{N_v} = \int d\vec{r}_v e^{-\beta U(\vec{r}_v)} \, .
  \label{eq:Z-pure-solvent}
\end{equation}

@@

The standard chemical potential of a solute in an ideally dilute solution is obtained from Eq.~(\ref{eq:chemical-potential-definition}) by setting the concentration to the standard concentration $C^\circ$, generally $1$ mol/L equivalent to $1/1,661$ particles/\AA$^{3}$, 
\begin{equation}
  \mu_u^\circ = -k_B T \ln \frac{q'_u}{C^\circ \Lambda_u^3} \, .
  \label{eq:chemical-potential-standard}
\end{equation}
The standard state of a dilute solution is defined as an ideally dilute solution at concentration $C^\circ$.~\cite{levine2009physicalchemistrybook6ed} Hence, Eq.~(\ref{eq:chemical-potential-standard}) is formally correct for real solutions as well.

@@

To derive an RBFE formulation from double-decoupling, we write expressions for the binding constants of $A$ and $B$ using Eq.~(\ref{eq:ddm-ratios-1}) and taking their ratio. While doing so, the standard state, pure solvent, and gas-phase terms cancel, and we obtain
\begin{equation}\label{eq:double-transformation-Kb-rbfe}
\frac{K_b(B)}{K_b(A)} = \frac{Z_{N_v,RB}}{Z_{N_v,RA}} \frac{Z_{N_v,A}}{Z_{N_v,B}}
\end{equation}
where the first ratio of partition functions involves receptor-ligand complexes and the second only the ligands in solution. Eq.~(\ref{eq:double-transformation-Kb-rbfe}), suggests that the relative binding free energy, $\Delta \Delta G_b$, of $B$ relative to $A$ can be expressed as the difference of the free energies of ``mutating'' ligand $A$ into ligand $B$ in the presence and absence of the receptor (Figure \ref{fig:fig_binding_thermodynamic_cycle}):
\begin{equation}\label{eq:double-transformation-dg-rbfe}
\Delta \Delta G_b = -k_B T \ln \frac{Z_{N_v,RB}}{Z_{N_v,RA}} - (-k_B T \ln \frac{Z_{N_v,B}}{Z_{N_v,A}}) = \Delta G_{\rm bound} - \Delta G_{\rm unbound}
\end{equation}
In practice, each ratio of partition functions must first be turned into an ensemble average to evaluate it. Similar to the preceding case, this step requires manipulating the integrals so that they are carried over the same set of degrees of freedom. As discussed in Section \ref{sec:relative-fe-protocol}, this step involves using alchemical topologies capable of describing both ligands. For example, among other features, alchemical topologies include atoms common to both ligands and dummy atoms that exist in one but not in the other ligand (Figure \ref{fig:fig_topology}). In one limit, called dual coordinate dual topology in Section \ref{sec:relative-fe-protocol}, all atoms of both ligands are present, and the partition functions in Eq.~(\ref{eq:double-transformation-dg-rbfe}) differ in which ligand is coupled to the system. A key aspect of constructing a suitable alchemical topology is the requirement that the additional degrees of freedom do not affect the RBFE estimate.
